---
title: "R for Data Science"
author: "sbsambado"
date: "4/12/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I am going through the 'R for Data Science' text by G. Grolemund and H. WIckham to find helpful code for students (https://r4ds.had.co.nz/index.html)

This book focuses on 'tidying' data and focuses on the packages `tidyverse` and `ggplot2` for data visualization.

helpful data frames to play around with 
- `mpg`
- `diamonds` 
- `nycflights13`
```{r}
library(tidyverse)
library(ggplot2)

```


ggplot template: adjust the <arguments> to your purposes

> ggplot(data = <DATA>) +
> <GEOM_FUNCTION>(mapping = aes(<MAPPING>)),
> stat = <STAT>,
> position = <POSITION>) +
> COORDINAT_FUNCTION> +
> <FACET_FUNCTION>

- aes(thetic) : visual property of the objects in your plot (i.e. size, shape, color of geom_function)

Common problems:
- every '(' is matched with a ')' ; i.e. no unclosed parenthesis/brackets
- every ' " ' is paired with another ' " '
- + in cosole means it is an uncomplete expression and it's waiting for you to finish it

```{r}
# examine of ggplot

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut)) +
  theme_classic()
```

4. Workflow: basics

Create new objects with '<-' , also '=' but that can get confusing later
object_name <- value

Object name must start with a letter, and can only contain letters, numbers, _ and .

Naming suggestions:
  i_use_snake_case
  otherPeopleUseCamelCase
  some.people.use.periods
- make is descriptive, but not long

4.3 Calling functions

function_name(arg1 = val1, arg2 = val2,...)

- if you can't remember the exact function spelling, start typing begining portion of the word and hit 'TAB' to see suggestions 
- if your function was successful it should pop up in the environment panel in upper right pane of R studio

5. Data transformation

Tibbles are data frames but flightly twesked to work better in tidyverse
Some abbreviations under column names
`int` stands for integers
`dbl` stands for doubles, or real numbers
`chr` stands for character vectors or strings
`dttm` stands for dat-times (a date + a time)

Types of variables
`lgl` stands for logical, vectors that contain only TRUE or FALSE
`fctr` stands for factors, which R uses to represent categorical variables with fixed possible values
`date` stands for dates


5.1.3 dplyr basics

Key dplyr functions that allow you to solve vast majority of data manipulation challenges
`filter()` pick observations by their values
`arrange()` reorder the rows
`select()` pick variables by their names
`mutate()` create new variables with functions of existing variables
`summarise()` collapse many values down to a single summary

These can all be used in conjunction with group_by() which changes the scope of each function from operating on the entire dataset to operating on it group by group. These 6 functions provide the verbs for a language of data manipulation

All verbs works similarly
  1. The first argument is a data frame
  2. The subsequent arguments describe what to do with the data frame, using the variable names (without quotes)
  3. The result is a new data frame
  
Together these properties make it easy to chain together multiple simple steps to achieve a complex results

5.2 Filter rows with filter

filter() allows you to subset observations based on their values

filter(argument1 = dataframe, argument2,n = expression that filter the dataframe)
```{r}
# nyc flight example
#library(tidyverse)
#library(nycflights13)

# if you want to save result, use assignment operator <-
jan1 <- filter(flights, month == 1, day == 2)

# want to save result, and print variable, wrap the assignment in parantheses
(jan1 <- filter(flights, month == 1, day == 2))
```

5.2.1 Comparisons

To filter effectively, you have to select observations that you want using the comparison operators
`>`
`<`
`>=`
`<=`
`==` equal
`!=` not equal

5.2.2 Logical operators

Multiple arguments to `filter()` are combined with "and" 
- every expression must be true in order for a row to be included in the output
- for combinations use Boolean operators
  `&`
  `|` or
  `!` not
```{r}
# flights that departed in november or december
filter(flights, month == 11 | month == 12)
```
Select every row where `x` is one of the values in `y` use ` x %in% y `
```{r}
nov_dec <- filter(flights, month %in% c(11,12))
```

5.2.3 Missing values
`NA` represents an unknown value
```{r}
# to determine if a value is missing, use `is.na()`
is.na(x)
```

filter() only includes rows wehre the condition is TRUE; it excludes FALSE and NA values
- if you want to preserve missing values, ask for them explicity
```{r}
# explicity ask for NA values

df <- tibble(x = c(1, NA, 3))
filter(df, is.na(x) | x > 1)
```

5.3 Arrange rows with arrange()

Works similar to filter(), except instead of selecting rows, it changes their order
- takes data frame and set of column names to order by
- missing values are always sorted at end

```{r}
arrange(flights, year, month, day)

# use desc() to re-order a column in descending order 

arrange(flights, desc(dep_delay))
```

5.4 select columns with select()

Want to narrow in on the 2134298 variables in your data set?
- use select()
- select() allows you to zoom in on useful subset using operations based on the names of variables
```{r}
# select columns by name
select(flights, year, month, day)

# select all columns between year and day (inclusive)
select(flights, year:day)

# select all columns except those from year to day (inclusive)
select(flights, -(year:day))
```

Helper functions that can be used within select
`starts_with("abc")`
`ends_with("abc")`
`contains("ijk")`
`matches("(.)\\1")`

rename() keeps all variables that aren't explicitly mentioned
```{r}
rename(flights, tail_num = tailnum)
```
Or use everything() helper in conjunction with select()
- helpful if you want to move variables to start of data frame
```{r}
select(flights, time_hour, air_time, everything())
```

5.5 Add new variables with mutate()
- add new columns that are functions of existing columns
- use View() to check if you added successfully

```{r}
flights_sml <- select(flights,
                      year: day,
                      ends_with('delay'),
                      distance, 
                      air_time)
mutate(flights_sml, 
       gain = dep_delay - arr_delay,
       speed = distance / air_time * 60)

mutate(flights_sml,
       gain = dep_delay - arr_delay,
       hours = air_time / 60,
       gain_per_hour = gain / hours)

# if you only want to keep new variables, use transmute()

transmute(flights, 
          gain = dep_delay - arr_delay,
          hours = air_time / 60,
          gain_per_hour = gain / hours)
```

5.5.1 Useful creation functions
mutate() can create new variables, but key property is that the function must be vectorized
- must take a vector of values as input, return a vector with the same number of values as output
  - arithmetic operators
      +, - , * , / , ^
  - modular arithmetic
    `%/%` integer divison 
    `%%` remiander
  - logs
    log(), log2(), log10()
  - offsets
    lead(), lag()
  - cumulative and rolling aggregates
    cumsum() : sum
    cumprod() : product
    cummin(), cummax() : min max
    cummean() : cumulative means
  - ranking
    min_rank(x)
    min_rank(desc(x))
    percent_rank(x)
    
    
5.6 Grouped summaries with summarise()

summarise() collapses a data frame into a single row
```{r}
summarise(flights, delay = mean(dep_delay, na.rm = TRUE))
```

5.6.1 Combining multiple operations with the pipe

3 steps to prepare data
  1. group flights by destination
  2. summarise to compute distance, average delay, and number of flights
  3. filter to remove noisy points and honolulu airport
```{r}
# want to explore relationship between the distance and average delay for each location

by_dest <- group_by(flights, dest)
delay <- summarise(by_dest,
                   count = n(),
                   dist = mean(distance, na.rm = TRUE),
                   delay = mean(arr_delay, na.rm = TRUE))

delay <- filter(delay, count > 20, dest !="HNL")

ggplot(data = delay, mapping = aes(x = dist, y = delay)) +
  geom_point(aes(size = count), alpha = 1/3) +
  geom_smooth(se = FALSE)

# or use the pipe %>%

delays  <- flights %>%
  group_by(dest) %>%
  summarise(
    count = n(),
    dist = mean(distance, na.rm = TRUE),
    delay = mean(arr_delay, na.rm = TRUE)
  ) %>%
  filter(count > 20, dest != 'HNL')
```

