---
title: "R for Data Science"
author: "sbsambado"
date: "4/12/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I am going through the 'R for Data Science' text by G. Grolemund and H. WIckham to find helpful code for students (https://r4ds.had.co.nz/index.html)

This book focuses on 'tidying' data and focuses on the packages `tidyverse` and `ggplot2` for data visualization.

helpful data frames to play around with 
- `mpg`
- `diamonds` 
- `nycflights13`
```{r}
library(tidyverse)
library(ggplot2)

```


ggplot template: adjust the <arguments> to your purposes

> ggplot(data = <DATA>) +
> <GEOM_FUNCTION>(mapping = aes(<MAPPING>)),
> stat = <STAT>,
> position = <POSITION>) +
> COORDINAT_FUNCTION> +
> <FACET_FUNCTION>

- aes(thetic) : visual property of the objects in your plot (i.e. size, shape, color of geom_function)

Common problems:
- every '(' is matched with a ')' ; i.e. no unclosed parenthesis/brackets
- every ' " ' is paired with another ' " '
- + in cosole means it is an uncomplete expression and it's waiting for you to finish it

```{r}
# examine of ggplot

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut)) +
  theme_classic()
```

4. Workflow: basics

Create new objects with '<-' , also '=' but that can get confusing later
object_name <- value

Object name must start with a letter, and can only contain letters, numbers, _ and .

Naming suggestions:
  i_use_snake_case
  otherPeopleUseCamelCase
  some.people.use.periods
- make is descriptive, but not long

4.3 Calling functions

function_name(arg1 = val1, arg2 = val2,...)

- if you can't remember the exact function spelling, start typing begining portion of the word and hit 'TAB' to see suggestions 
- if your function was successful it should pop up in the environment panel in upper right pane of R studio

5. Data transformation

Tibbles are data frames but flightly twesked to work better in tidyverse
Some abbreviations under column names
`int` stands for integers
`dbl` stands for doubles, or real numbers
`chr` stands for character vectors or strings
`dttm` stands for dat-times (a date + a time)

Types of variables
`lgl` stands for logical, vectors that contain only TRUE or FALSE
`fctr` stands for factors, which R uses to represent categorical variables with fixed possible values
`date` stands for dates


5.1.3 dplyr basics

Key dplyr functions that allow you to solve vast majority of data manipulation challenges
`filter()` pick observations by their values
`arrange()` reorder the rows
`select()` pick variables by their names
`mutate()` create new variables with functions of existing variables
`summarise()` collapse many values down to a single summary

These can all be used in conjunction with group_by() which changes the scope of each function from operating on the entire dataset to operating on it group by group. These 6 functions provide the verbs for a language of data manipulation

All verbs works similarly
  1. The first argument is a data frame
  2. The subsequent arguments describe what to do with the data frame, using the variable names (without quotes)
  3. The result is a new data frame
  
Together these properties make it easy to chain together multiple simple steps to achieve a complex results

5.2 Filter rows with filter

filter() allows you to subset observations based on their values

filter(argument1 = dataframe, argument2,n = expression that filter the dataframe)
```{r}
# nyc flight example
#library(tidyverse)
#library(nycflights13)

# if you want to save result, use assignment operator <-
jan1 <- filter(flights, month == 1, day == 2)

# want to save result, and print variable, wrap the assignment in parantheses
(jan1 <- filter(flights, month == 1, day == 2))
```

5.2.1 Comparisons

To filter effectively, you have to select observations that you want using the comparison operators
`>`
`<`
`>=`
`<=`
`==` equal
`!=` not equal

5.2.2 Logical operators

Multiple arguments to `filter()` are combined with "and" 
- every expression must be true in order for a row to be included in the output
- for combinations use Boolean operators
  `&`
  `|` or
  `!` not
```{r}
# flights that departed in november or december
filter(flights, month == 11 | month == 12)
```
Select every row where `x` is one of the values in `y` use ` x %in% y `
```{r}
nov_dec <- filter(flights, month %in% c(11,12))
```

5.2.3 Missing values
`NA` represents an unknown value
```{r}
# to determine if a value is missing, use `is.na()`
is.na(x)
```

filter() only includes rows wehre the condition is TRUE; it excludes FALSE and NA values
- if you want to preserve missing values, ask for them explicity
```{r}
# explicity ask for NA values

df <- tibble(x = c(1, NA, 3))
filter(df, is.na(x) | x > 1)
```

5.3 Arrange rows with arrange()

Works similar to filter(), except instead of selecting rows, it changes their order
- takes data frame and set of column names to order by
- missing values are always sorted at end

```{r}
arrange(flights, year, month, day)

# use desc() to re-order a column in descending order 

arrange(flights, desc(dep_delay))
```

5.4 select columns with select()

Want to narrow in on the 2134298 variables in your data set?
- use select()
- select() allows you to zoom in on useful subset using operations based on the names of variables
```{r}
# select columns by name
select(flights, year, month, day)

# select all columns between year and day (inclusive)
select(flights, year:day)

# select all columns except those from year to day (inclusive)
select(flights, -(year:day))
```

Helper functions that can be used within select
`starts_with("abc")`
`ends_with("abc")`
`contains("ijk")`
`matches("(.)\\1")`

rename() keeps all variables that aren't explicitly mentioned
```{r}
rename(flights, tail_num = tailnum)
```
Or use everything() helper in conjunction with select()
- helpful if you want to move variables to start of data frame
```{r}
select(flights, time_hour, air_time, everything())
```

5.5 Add new variables with mutate()
- add new columns that are functions of existing columns
- use View() to check if you added successfully

```{r}
flights_sml <- select(flights,
                      year: day,
                      ends_with('delay'),
                      distance, 
                      air_time)
mutate(flights_sml, 
       gain = dep_delay - arr_delay,
       speed = distance / air_time * 60)

mutate(flights_sml,
       gain = dep_delay - arr_delay,
       hours = air_time / 60,
       gain_per_hour = gain / hours)

# if you only want to keep new variables, use transmute()

transmute(flights, 
          gain = dep_delay - arr_delay,
          hours = air_time / 60,
          gain_per_hour = gain / hours)
```

5.5.1 Useful creation functions
mutate() can create new variables, but key property is that the function must be vectorized
- must take a vector of values as input, return a vector with the same number of values as output
  - arithmetic operators
      +, - , * , / , ^
  - modular arithmetic
    `%/%` integer divison 
    `%%` remiander
  - logs
    log(), log2(), log10()
  - offsets
    lead(), lag()
  - cumulative and rolling aggregates
    cumsum() : sum
    cumprod() : product
    cummin(), cummax() : min max
    cummean() : cumulative means
  - ranking
    min_rank(x)
    min_rank(desc(x))
    percent_rank(x)
    
    
5.6 Grouped summaries with summarise()

summarise() collapses a data frame into a single row
```{r}
summarise(flights, delay = mean(dep_delay, na.rm = TRUE))
```

5.6.1 Combining multiple operations with the pipe

3 steps to prepare data
  1. group flights by destination
  2. summarise to compute distance, average delay, and number of flights
  3. filter to remove noisy points and honolulu airport
```{r}
# want to explore relationship between the distance and average delay for each location

by_dest <- group_by(flights, dest)
delay <- summarise(by_dest,
                   count = n(),
                   dist = mean(distance, na.rm = TRUE),
                   delay = mean(arr_delay, na.rm = TRUE))

delay <- filter(delay, count > 20, dest !="HNL")

ggplot(data = delay, mapping = aes(x = dist, y = delay)) +
  geom_point(aes(size = count), alpha = 1/3) +
  geom_smooth(se = FALSE)

# or use the pipe %>%

delays  <- flights %>%
  group_by(dest) %>%
  summarise(
    count = n(),
    dist = mean(distance, na.rm = TRUE),
    delay = mean(arr_delay, na.rm = TRUE)
  ) %>%
  filter(count > 20, dest != 'HNL')
```

5.6.3 Counts

While performing an aggregation, good idea toinclude a count (n()) or a count of non-missing values (sum(!is.na(x)))
```{r}
not_cancelled <- flights %>%
  filter(!is.na(dep_delay), !is.na(arr_delay))

rename(flights, tail_num = tailnum)
delays <- not_cancelled %>%
  group_by(tail_num) %>%
  summarise(
    delay = mean(arr_delay)
  )

ggplot(data = delays, mapping = aes(x = delay)) +
  geom_freqpoly(binwidth = 10)

# make a scatter plot
ggplot(data = delays, mapping = aes(x = n, y = delay)) +
  geom_point(alpha = 1/10)

# want to filter to make map prettier
delays %>%
  filter(n > 25) %>%
  ggplot(mapping = aes(x = n, y = delay)) +
  geom_point(alpha = 1/10)
```

Rstudio tip: keyboard shortcut Cmd + Shift + P resends previous sent chunk from editor to console


Look at average performance of batters in baseball is related to number of itmes they're at bat
```{r}
install.packages('Lahman')
library(Lahman)
batting <- as_tibble(Lahman::Batting) #Lahman package

batters <- batting %>%
  group_by(playerID) %>%
  summarise(
    ba = sum(H, na.rm = TRUE) /sum(AB, na.rm = TRUE),
    ab = sum(AB, na.rm = TRUE)
  )

batters %>%
  filter(ab > 100) %>%
  ggplot(mapping = aes(x = ab, y = ba)) +
  geom_point() +
  geom_smooth(se = FALSE) # using gam method y ~ s(x, bs = 'cs')
```

5.6.4 Useful summary functions

*Measures of location*
mean(x) : sum divided by length
meadian(x) : where 50% of x is below or above that value
```{r}
not_cancelled %>%
  group_by(year, month, day) %>%
  summarise(
    avg_delay1 = mean(arr_delay),
    avg_delay2 = mean(arr_delay[arr_delay > 0]) # average positive delay
  )
```

*Measures of spread*
sd(x)
IQR(x)
mad(x)

```{r}
not_cancelled %>%
  group_by(dest) %>%
  summarise(distance_sd = sd(distance)) %>%
  arrange(desc(distance_sd))
```
*Measures of rank*
min(x)
quantile(x, 0.25)
max(x)
```{r}
# when do the first and last flights leave each day?
not_cancelled %>%
  group_by(year, month, day) %>%
  summarise(
    first = min(dep_time),
    last = max(dep_time)
  )
```

*Measures of position*
first(x)
nth(x)
last(x)
```{r}
not_cancelled %>%
  group_by(year, month, day) %>%
  summarise(
    first_dep = first(dep_time),
    last_dep = last(dep_time)
  )

# want to filter to give each observation a separate row
not_cancelled %>%
  group_by(year, month, day) %>%
  mutate(r = min_rank(desc(dep_time))) %>%
  filter(r %in% range(r))
```

*Counts*
n(): takes no arguments and returns size of current group
sum(!is.na(x)) : count number of non-missing values
n_distinct(x) : count number of distinct values
```{r}
not_cancelled %>%
  group_by(dest) %>%
  summarise(carriers = n_distinct(carrier)) %>%
  arrange(desc(carriers))

# dplyr uses simple helper
not_cancelled %>%
  count(dest)

# provide weight variable
not_cancelled %>%
  count(tailnum, wt = distance)


```
Counts and proportions of logical values
-when used with  numeric functions
  TRUE = 1
  FALSE = 0
  - makes sum() and mean() useful
  - sum(x) gives number of TRUEs in x
  - mean(x) gives proprtion
```{r}
# how many flights leave before 5 AM?
not_cancelled %>%
  group_by(year, month, day) %>%
  summarise(n_early = sum(dep_time < 500))

# what proportion of flights are delayed by more than an hour?
not_cancelled %>%
  group_by(year, month, day) %>%
  summarise(hour_prop = mean(arr_delay > 60))
```

5.6.5 Grouping by multiple variables

group by multiple variables, each summary peels off one level of grouping
- it's okay for sums and counts, but think about weighting means and variances
```{r}

daily <- group_by(flights, year, month, day)
(per_day <- summarise(daily, flights = n()))
(per_month <- summarise(per_day, flights = sum(flights)))
(per_year <- summarise(per_month, flights = sum(flights)))

```
5.6.6 Ungrouping
```{r}
daily %>%
  ungroup() %>%
  summarise(flights = n())
```

5.7 Grouped mutates (and filters)

grouping is most useful in conjunction with summarise(), but you can also use mutate() and filter()
```{r}
# find the worst members of each group

flights_sml %>%
  group_by(year, month, day) %>%
  filter(rank(desc(arr_delay)) < 10)

# find all groups bigger than a threshold

popular_dests <- flights %>%
  group_by(dest) %>%
  filter(n() > 365)
popular_dests

# standardize to compute per group metrics

popular_dests %>%
  filter(arr_delay > 0) %>%
  mutate(prop_delay = arr_delay / sum(arr_delay)) %>%
  select(year:day, dest, arr_delay, prop_delay)
``` 
A grouped filter is grouped mutate followed by an ungrouped filter. 

Functions that work most naturally in grouped mutates and filters are known as window functions

Chapter 7 Exploratory Data Analysis (EDA)

7.1 Introduction

EDA is an iterative cycle
1. generate questions about your data
2. search for answers by visualizing, transforming, and modelling your data
3. use what you learn to refine your questions and/or generate new questions

Feel free to investigate every idea that occurs to you!

7.1. prerequisities
`tidyverse` which contains dplyr

7.2 Questions

1. What type of variation occurs within my variables?
2. What type of covariation occurs between my variables?