---
title: "demystifying stat_layers in ggplot2"
author: "sbsambado"
date: "2/15/2021"
output: html_document
---

```{r setup, include=FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
```

just taking code from June Choe
https://yjunechoe.github.io/posts/2020-09-26-demystifying-stat-layers-ggplot2/

Thought it was helpful to think about stat_layers vs geom_layers

```{r message = FALSE}

# make data
simple_data <- tibble(group = factor(rep(c("A", "B"), each = 15)),
                      subject = 1:30,
                      score = c(rnorm(15, 40, 20), rnorm(15, 60, 10)))

```

```{r the long way}
simple_data_bar <- simple_data %>%
  group_by(group) %>% 
  summarize(
    mean_score = mean(score),
    .groups = 'drop'
  )
  
simple_data_errorbar <- simple_data %>% 
  group_by(group) %>% 
  summarize(
    mean_score = mean(score),
    se = sqrt(var(score)/length(score)),
    .groups = 'drop'
  ) %>% 
  mutate(
    lower = mean_score - se,
    upper = mean_score + se
  )

ggplot() +
  geom_col(
    aes(x = group, y = mean_score),
    data = simple_data_bar
  ) +
  geom_errorbar(
    aes(x = group, y = mean_score, ymin = lower, ymax = upper),
    data = simple_data_errorbar
  )
```

```{r the short way}
simple_data %>% 
  ggplot(aes(group, score)) +
  stat_summary(geom = "bar") +
  stat_summary(geom = "errorbar")
```
Let’s analyze stat_summary() as a case study to understand how stat_*()s work more generally. I think that stat_summary() is a good choice because it’s a more primitive version of many other stat_*()s and is likely to be the one that you’d end up using the most for visualizations in data science.

```{r }
height_df <- tibble(group = "A",
                    height = rnorm(30, 170, 10))

# geom way
height_df %>% 
  ggplot(aes(x = group, y = height)) +
  geom_point()

# stat way
height_df %>% 
  ggplot(aes(x = group, y = height)) +
  stat_summary()
```
First, we see from the documentation of stat_summary() that this mean_se() thing is the default value for the fun.data argument (we’ll talk more on this later).

Next, let’s call it in the console to see what it is:
type `mean_se` in console

function (x, mult = 1) 
  {
      x <- stats::na.omit(x)
      se <- mult * sqrt(stats::var(x)/length(x))
      mean <- mean(x)
      new_data_frame(list(y = mean, ymin = mean - se, ymax = mean + 
          se), n = 1)
  }
  
  
This is actually really important: stat_summary() summarizes one dimension of the data. 11 mean_se() threw an error when we passed it our whole data because it was expecting just a vector of the variable to be summarized.
```{r}
# wrong
# mean_se(height_df)

# right
mean_se(height_df$height)

```

You know how else we can check that this is the case? With this neat function called layer_data().

We can pull the data that was used to draw the pointrange by passing our plot object to layer_data() and setting the second argument to 1 12 :

```{r}
pointrange_plot <- height_df %>% 
  ggplot(aes(x = group, y = height)) +
  stat_summary()

layer_data(pointrange_plot, 1) # see all the info
```

layer_data(pointrange_plot, 1)
  

### Interim Summary 2
To summarize this section (ha!), stat_summary() works in the following order:

The data that is passed into ggplot() is inherited if one is not provided

The function passed into the fun.data argument applies transformations to (a part of) that data (defaults to mean_se())

The result is passed into the geom provided in the geom argument (defaults to pointrange).

If the data contains all the required mapppings for the geom, the geom will be plotted.

And to make things extra clear & to make stat_summary() less mysterious, we can explicitly spell out the two arguments fun.data and geom that we went over in this section.
```{r}
height_df %>% 
  ggplot(aes(x = group, y = height)) +
  stat_summary(
    geom = "pointrange",
    fun.data = mean_se
  )
```

### Putting STAT to use

```{r message = FALSE}
remotes::install_github("allisonhorst/palmerpenguins")
data("penguins", package = "palmerpenguins")
my_penguins <- na.omit(penguins) # get rid of NAs
```


```{r}
my_penguins %>% 
  ggplot(aes(sex, body_mass_g)) +
  stat_summary(
    fun.data = ~mean_se(., mult = 1.96), # Increase `mult` value for bigger interval!
    geom = "errorbar",
  )

bind_rows(
  mean_se(my_penguins$body_mass_g[my_penguins$sex == "female"], mult = 1.96),
  mean_se(my_penguins$body_mass_g[my_penguins$sex == "male"], mult = 1.96),
)
```
A color-coded bar plot of medians
```{r}
calc_median_and_color <- function(x, threshold = 40) {
  tibble(y = median(x)) %>% 
    mutate(fill = ifelse(y < threshold, "pink", "grey35"))
}

my_penguins %>% 
  ggplot(aes(species, bill_length_mm)) +
  stat_summary(
    fun.data = calc_median_and_color,
    geom = "bar"
  )
```

Pointrange plot with changing size
```{r}
my_penguins %>% 
  ggplot(aes(species, bill_depth_mm)) +
  stat_summary(
    fun.data = function(x) {
      
      scaled_size <- length(x)/nrow(my_penguins)
      
      mean_se(x) %>% 
        mutate(size = scaled_size)
    }
  )
```

