---
title: "YaRrr! A pirates guide to R"
author: "sbsambado"
date: "4/5/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

YaRrr! The Pirate's Guide to R
Nathaniel D Phillips
2018-01-22

Thought I'd run through this to see if I can find helpful tidbits of code for the undergradutes and students in Biometry 146.

Chapter 2
Debugging in R - share this with class

'When you are programming, you will *always*, and I do mean always, make errors (also called bugs) in your code. You might misspell a function, include an extra comma, or some days…R just won’t want to work with you (again, see section Why R is like a Relationship)' lol
https://bookdown.org/ndphillips/YaRrr/debugging.html

```{r}
# install data set used in this book 
#install.packages('yarrr')
#library(yarrr)

# exploring data
head(pirates)
names(pirates)

```

3.2 Descriptive statistics
```{r}
mean(pirates$age)
max(pirates$height)

# how many pirates are there of each sex
table(pirates$sex)

# calculate statistics for different groups of pirates

# calculate mean age, for each sex
aggregate(formula = age ~ sex, data = pirates, FUN = mean)

```

3.3 Plotting 
```{r}
plot(x = pirates$height,
     y = pirates$weight,
     main = 'scatterplot of pirate data!',
     xlab = 'height (cm)',
     ylab = 'weight (kg)',
     pch = 16,
     col = gray(0.0, 0.1))
grid() # add gridlines

# create linear regression model
model <- lm(formula = weight ~ height,
            data = pirates)
abline(model, col = 'blue') # add regression to plot

# scatterplots are great for showing relationship between two continuous variables
```
3.4 Hypothesis tests
```{r}
# age by headband t-test

t.test(formula = age ~ headband,
       data = pirates, 
       alternative = 'two.sided')

# significant correlation between a pirate's height and weight?
cor.test(formula = ~ height + weight,
         data = pirates)
# ^ yes, pvalue is 2.2e-16

# ANOVA testing between number of tattoos pirates have based on their favorite sword
tat.sword.lm <- lm(formula = tattoos ~ sword.type, 
                   data = pirates)
anova(tat.sword.lm)
```

3.5 Regression analysis
```{r}
# regression analysis to see if pirate's age, weight, tatoos predicts how many treasure chests they've found

tchests.model <- lm(formula = tchests ~ age + weight + tattoos, data = pirates)

summary(tchests.model)
```

3.6 Bayesian statistics
```{r}
#install.packages('BayesFactor')
#library(BayesFactor)

# bayesian version of t-tests

ttestBF(formula = age ~ headband,
        data = pirates)

# we got bayes factor of 0.12, which is strong evidence for the null hypothesis (mean age does not differ between pirates with or without headbands)
```

4.4 Objects and functions
Almost everything in R is either an object or function
object <- function()

object = thing, like number, dataset; different shapes and sizes

function = procedure that takes one or more objects are arguments (aka inputs), does someting with those objects, then returns a new object

4.4.1 Numbers vs characters
numbers = 1, 10, 3.14
charactes = 'madisen', 'brian', '10'


5.1 Scalers
scalar: an object that is just a single value like a number or a name. 
```{r}
# numeric scalers, you can do basic arithmetic operations
a <- 100
b <- 3 / 100
c <- (a + b) / b

# character scales aka STRINGS
d <- 'ship'
e <- 'cannon'

```

5.2 Vectors
vector: object that is a combination of several scalars stored as a single object

Vectors contain either numbers or characters, not both
```{r}
# different ways to create vectors

# numeric vectors
one <- c(1,5,9)
two <- 1:9
three <- seq(0,6,2)
four <- rep(c(7,8), times = 2, each = 2, length.out = 90)
print(four) # length.out is desired length of the final sequence

# character vectors
char.vec <- c('ceci', 'nest', 'pas', 'une', 'pipe')
char.vec
```

